---
title: "Analysis of U.S. Residential Mortgages from 1998-2021"
author: "Kendall Beaver"
#date: "2024-02-17"
output: pdf_document
header-includes:
  - \usepackage{caption}
  - \usepackage{float}
  - \usepackage{placeins}
  - \captionsetup[table]{justification=centering}
---

# Overview

This report analyzed over 20 years of U.S. residential mortgages (1998-2021) that were gathered from the [National Mortgage Database](https://www.fhfa.gov/PolicyProgramsResearch/Programs/Pages/National-Mortgage-Database.aspx) (NMDB®). Most notable is the fact that the dominant number of borrowers has reversed from two borrowers to a single borrower per mortgage, and recent regulation has caused lenders to approve mortgages for applicants with a good or excellent credit score more than ever before. This report assesses how interest rates have affected mortgage demand over two decades; how regulations have affected borrower creditworthiness; demographic changes in age, gender, and number of mortgage signers; and finally what the typical borrower profile looks like currently.

# I. How Interest Rates Have Affected Mortgage Demand

Contrary to popular belief that the Federal Reserve sets mortgage interest rates, lenders specifically set interest rates based upon market conditions. Figure 1 shows the annual average interest rate from 1998–2021, which began at 7.1% in 1998, hit a peak of 8.4% in 2000, then fluctuated as it descended over the next twenty years to an all-time low of 3.0% in 2021.

```{r 1, tidy=TRUE, message=FALSE, warning=FALSE, results='hide', echo=FALSE, fig.align='center'}

if (!require(pacman))
  install.packages("pacman")

pacman::p_load(dplyr, ggplot2, gridExtra, lubridate, reshape2, scales, stringr, forcats, kableExtra, ggrepel)

library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
library(lubridate)
library(reshape2)
library(scales)
library(stringr)
library(tidyr)
library(forcats)
library(kableExtra)
library(ggrepel)

mortgages <- read.csv("../data/mortgages.csv")

# Let's make the columns more human readable/friendly
mortgages <- mortgages %>%
  rename(SERIES_ID = SERIESID,
         GEO_LEVEL = GEOLEVEL,
         GEO_ID = GEOID,
         GEO_NAME = GEONAME)

unique(mortgages$GEO_LEVEL)
# National
# Census Region
# Census Division
# State

# Let's clean this up and turn the numeric values into numbers I can understand
mortgages %>%
  mutate(RELATIVE_VALUE = paste0(gsub("\\D", "", VALUE), "%"))

mortgages <- mortgages %>%
  filter(GEO_LEVEL == "National") %>%
  mutate(RELATIVE_VALUE = case_when(
    str_detect(SERIES_ID, "TOT_ORIG") ~ paste0(comma(VALUE * 1000)),
    str_detect(SERIES_ID, "AVE_LOANAMT") |
      str_detect(SERIES_ID, "AVE_PROPVAL") ~ paste0("$", comma(VALUE * 1000)),
    str_detect(SERIES_ID, "PCT") |
      str_detect(SERIES_ID, "DTI") |
      str_detect(SERIES_ID, "RATE") |
      str_detect(SERIES_ID, "LTV") |
      str_detect(SERIES_ID, "AVE_TERM") ~ paste0(format(VALUE, nsmall = ifelse(floor(VALUE) == VALUE, 0, 1)), "%"),
    str_detect(SERIES_ID, "AVE_AGE_BORROWER") ~ paste0(round(VALUE), " years old"),
    str_detect(SERIES_ID, "AVE_VANTAGESCR") ~ paste0(VALUE, " credit score"),
    TRUE ~ as.character(VALUE)))
```

```{r interest_plot, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', out.width="65%", fig.cap="Annual average mortgage interest rate from 1998–2021.", fig.pos='H'}

# Data for interest rates
interest_rates <- mortgages %>% 
  filter(MARKET == "All Mortgages",
         SERIES_ID == "AVE_INTRATE")

p_interest <- ggplot(interest_rates, aes(x = YEAR, y = VALUE)) +
  geom_line(color = "red", linewidth = 1) +
  labs(
    x = "\nYear",
    y = "Average Interest Rate\n"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal()

print(p_interest)

#grid::grid.text(
#  "Figure 1. Annual average interest rate from 1998–2021.",
#  y = unit(-0.1, "npc"),
#  gp = grid::gpar(fontsize = 11)
#)
```

Interest rates typically have an inverse relationship with borrower demand, where low interest rates typically increases the demand for mortgages, while high rates suppresses demand. However, the greatest amount of mortgage originations, 20,574,000, occurred in 2003 when rates were not at their lowest, 5.7% (Table 1). The lowest rate, as already mentioned, was 3.0% in 2021, which resulted in 13,361,000 mortgage originations. Therefore, the relationship between interest rates and demand is a generic relationship, not a definitive one.

```{r new_table_1, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE}

mortgage_table_data <- mortgages %>%
  filter(MARKET %in% c("All Mortgages (Home Purchase)",
                       "All Mortgages (Refinance)"),
         SERIES_ID == "TOT_ORIG") %>%
  mutate(
    Mortgage_Type = ifelse(grepl("Home Purchase", MARKET),
                           "Home Purchase", "Refinance"),
    Count = VALUE * 1000
  ) %>%
  select(YEAR, Mortgage_Type, Count)

mortgage_table_data <- mortgage_table_data %>%
  group_by(YEAR) %>%
  mutate(Percent = Count / sum(Count)) %>%
  ungroup()

interest_rates <- mortgages %>%
  filter(SERIES_ID == "AVE_INTRATE",
         MARKET == "All Mortgages") %>%
  select(YEAR, VALUE, RELATIVE_VALUE)

interest_extremes <- interest_rates %>%
  summarise(
    Min_Year = YEAR[which.min(VALUE)],
    Max_Year = YEAR[which.max(VALUE)]
  )

min_year <- interest_extremes$Min_Year
max_year <- interest_extremes$Max_Year

total_mortgages <- mortgage_table_data %>%
  group_by(YEAR) %>%
  summarise(Total = sum(Count), .groups = "drop")

total_extremes <- total_mortgages %>%
  summarise(
    Total_Min_Year = YEAR[which.min(Total)],
    Total_Max_Year = YEAR[which.max(Total)]
  )

total_min_year <- total_extremes$Total_Min_Year
total_max_year <- total_extremes$Total_Max_Year

years_to_include <- unique(c(
  1998, 2021, 2003, 2007, 2008, 2009,
  min_year, max_year, total_min_year, total_max_year, 2019, 2020
))

final_table <- mortgage_table_data %>%
  filter(YEAR %in% years_to_include) %>%
  left_join(interest_rates %>% select(YEAR, RELATIVE_VALUE), by = "YEAR") %>%
  pivot_wider(
    names_from = Mortgage_Type,
    values_from = c(Count, Percent)
  ) %>%
  mutate(
    Total_Mortgages = `Count_Home Purchase` + `Count_Refinance`,
    `Average Interest Rate` = RELATIVE_VALUE,
    `All Mortgages (Count)` = comma(Total_Mortgages),
    `New Home Mortgages (Count)` = comma(`Count_Home Purchase`),
    `Refinance Mortgages (Count)` = comma(`Count_Refinance`),
    `New Home Mortgage (%)` = percent(`Percent_Home Purchase`, accuracy = 0.1),
    `Refinance Mortgage (%)` = percent(`Percent_Refinance`, accuracy = 0.1)
  ) %>%
  select(
    YEAR,
    `Average Interest Rate`,
    `All Mortgages (Count)`,
    `New Home Mortgages (Count)`,
    `Refinance Mortgages (Count)`,
    `New Home Mortgage (%)`,
    `Refinance Mortgage (%)`
  ) %>%
  arrange(YEAR)

colnames(final_table)[1] <- "Year"

knitr::kable(
  final_table,
  caption = "Selected years of interest rates and mortgages by count, type, and percentage.",
  align = "ccccccc"
)
```

There are two types of mortgages that make up residential mortgages: new home vs. refinance. While there were roughly an equal amount of refinance and new home mortgages in 1999, 54.4% vs. 45.6%, by the end of 2021 there were significantly more refinance than new home mortgages, 63.5% vs. 36.5%. Figure 2 shows how the demand for each mortgage has fluctuated over time, but what has been a more consistent trend is that when interest rates drop, there is greater demand for refinance mortgages than for new home mortgages as seen from 2000 to 2003, 2008 to 2009, and 2019 to 2020 (see Table 2 for details).

```{r 2, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=6, fig.height=3.8, fig.cap="Stacked area chart of home purchase and refinance mortgages from 1998-2021.", fig.pos='H'}

#{r 2, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=5.5, fig.height=3.8}

### SUPPLEMENTAL STACKED AREA PLOT: "Home Purchase" and "Refinance"

# Filter data for home purchases and refinances
home_purchases_vs_refinances <- mortgages %>% 
  filter((MARKET == "All Mortgages (Home Purchase)" | 
            MARKET == "All Mortgages (Refinance)") & 
           (SERIES_ID == "TOT_ORIG"))

# Rename MARKET values so they appear clean in the legend
home_purchases_vs_refinances$MARKET[grepl("Home Purchase", home_purchases_vs_refinances$MARKET)] <- "Home Purchase"
home_purchases_vs_refinances$MARKET[grepl("Refinance", home_purchases_vs_refinances$MARKET)] <- "Refinance"

# Stacked area plot
p3_home_purchases <- ggplot(home_purchases_vs_refinances, aes(x = YEAR, y = VALUE, fill = MARKET)) +
  geom_area(stat = "identity", alpha = 1, col = "black", linewidth = 0.5) +
  labs(
    x = "\nYear",
    y = "Count of Mortgages Originated\n",
    fill = "Mortgage Type"
  ) +
  scale_fill_manual(values = c("#b2df8a", "#33a02c")) +
  scale_y_continuous(labels = function(x) label_comma(accuracy = 1)(x * 1000)) +
  theme_minimal()

print(p3_home_purchases)

ggsave("../plots/stacked_area_chart_refinance_new_home_mortgages.png", width = 6, height = 3.8, dpi = 300)

# Caption 2. Stacked Area Chart of Total Yearly Count of New Mortgage Originations by Mortgage Type showing that over 20 years, mortgage borrowers have always been originating refinance mortgages than home purchases. (Then include table of values of 3 variables.)
```

# II. How Regulations Have Affected the Creditworthiness of Borrowers

Prior to the subprime mortgage crisis that began in 2007 and ended in 2009, mortgage lenders were more willing and able to approve loans for applicants who had a low credit rating of Very Poor (300-499), Poor (500-600), or Fair (601-660), where the likelihood of default increases as the credit rating decreases. Table 2 highlights the years (in a green color) when each credit rating experienced their greatest proportion of mortgage originations.

```{r credit_score_table, echo=FALSE, fig.pos='H'}

credit_score_table <- mortgages %>%
  filter(
    MARKET == "All Mortgages",
    SERIES_ID %in% c(
      "PCT_VS_EXCELLENT",
      "PCT_VS_GOOD",
      "PCT_VS_FAIR",
      "PCT_VS_POOR",
      "PCT_VS_VERYPOOR"
    ),
    YEAR %in% c(1998, 1999, 2000, 2005, 2008, 2009, 2010, 2011, 2012, 2014, 2015, 2016, 2017, 2021)
  ) %>%
  mutate(
    Credit_Score = case_when(
      SERIES_ID == "PCT_VS_VERYPOOR" ~ "VeryPoor",
      SERIES_ID == "PCT_VS_POOR" ~ "Poor",
      SERIES_ID == "PCT_VS_FAIR" ~ "Fair",
      SERIES_ID == "PCT_VS_GOOD" ~ "Good",
      SERIES_ID == "PCT_VS_EXCELLENT" ~ "Excellent"
    ),
    Percent_num = as.numeric(gsub("%", "", as.character(VALUE))),
    Percent = paste0(Percent_num, "%")
  ) %>%
  select(YEAR, Credit_Score, Percent_num, Percent) %>%
  pivot_wider(
    names_from = Credit_Score,
    values_from = c(Percent_num, Percent)
  ) %>%
  arrange(YEAR)

credit_score_table$Percent_VeryPoor <- cell_spec(
  credit_score_table$Percent_VeryPoor,
  format = "latex",
  background = ifelse(
    credit_score_table$YEAR == 1999 & credit_score_table$Percent_num_VeryPoor == 3.4,
    "#d4edda", "white"
  )
)

credit_score_table$Percent_Poor <- cell_spec(
  credit_score_table$Percent_Poor,
  format = "latex",
  background = ifelse(
    credit_score_table$YEAR == 2000 & credit_score_table$Percent_num_Poor == 14.8,
    "#d4edda", "white"
  )
)

credit_score_table$Percent_Fair <- cell_spec(
  credit_score_table$Percent_Fair,
  format = "latex",
  background = ifelse(
    credit_score_table$YEAR == 2005 & credit_score_table$Percent_num_Fair == 22.5,
    "#d4edda", "white"
  )
)

credit_score_table$Percent_Good <- cell_spec(
  credit_score_table$Percent_Good,
  format = "latex",
  background = ifelse(
    credit_score_table$YEAR %in% c(2014, 2017) & credit_score_table$Percent_num_Good == 56.6,
    "#d4edda", "white"
  )
)

credit_score_table$Percent_Excellent <- cell_spec(
  credit_score_table$Percent_Excellent,
  format = "latex",
  background = ifelse(
    credit_score_table$YEAR == 2012 & credit_score_table$Percent_num_Excellent == 37.5,
    "#d4edda", "white"
  )
)

display_table <- credit_score_table %>%
  select(
    YEAR,
    Percent_VeryPoor,
    Percent_Poor,
    Percent_Fair,
    Percent_Good,
    Percent_Excellent
  )

kbl(
  display_table,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  align = "c",
  col.names = c(
    "Year",
    "(300–499)",
    "(500–600)",
    "(601–660)",
    "(661–780)",
    "(781–850)"
  ),
  caption = "Selected years of mortgage originations based on proportion of credit scores."
) %>%
  add_header_above(c(
    " " = 1,
    "Very Poor" = 1,
    "Poor" = 1,
    "Fair" = 1,
    "Good" = 1,
    "Excellent" = 1
  )) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE
  )
```

In 1999 and 2000, Very Poor- and Poor-rated mortgages had the highest shares of originations at 3.4% and 14.8%, respectively, which occurred before an eight-month-long recession in 2001 (March-November), and over twenty years later, both ratings became almost non-existent: 0.1% and 1.9%, respectively. Fair-rated mortgages had their highest share of originations in 2005, at 22.5%, then fluctuated as it decreased to 9.1% at the end of 2021, a drop of 59.6% since its pinnacle.

Then in 2008, the major investment bank Bear Stearns collapsed in March, followed by Lehman Brothers in September. The failure of Lehman Brothers triggered the collapse of the global financial system, and over the following year, there were significantly less Poor- and Fair-rated mortgage originations (-50.7% and -31.9%, respectively) and more Excellent-rated mortgages (43.5%), while Good-rated mortgages remained about the same (-1.0%).

On July 21, 2010, the U.S. government enacted the Dodd-Frank Wall Street Reform and Consumer Protection Act ("Dodd-Frank"), to reduce investment in risky mortgage-backed securities and prevent another housing financial crisis. In response, lenders originated more Excellent-rated mortgages than ever before, 36.8% between 2010 and 2011, and 37.5% in 2012, the highest proportion of origination for this credit rating.

Then on January 10, 2014, Dodd-Frank imposed the Ability-to-Repay/Qualified Mortgage (ATR/QM) rule which required a lender to issue a mortgage based on sound principles—more specifically, a lender must make a "reasonable and good faith determination" that a borrower can repay a mortgage before issuing it. By the end of the year, Good-rated mortgages reached their pinnacle, 56.6%, and remained around the same percentage since then, reaching 56.6% again in 2017 and ending at 55.5% in 2021.

```{r credit_score_plot_new, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=7, fig.height=3.8, fig.cap="Stacked area chart of borrower credit scores based on percentage from 1998–2021.", fig.pos='H'}

# STACKED AREA PLOT: Credit score

mortgages_credit_score <- mortgages %>%
  filter(MARKET == "All Mortgages",
         SERIES_ID %in% c("PCT_VS_VERYPOOR",
                          "PCT_VS_POOR",
                          "PCT_VS_FAIR",
                          "PCT_VS_GOOD",
                          "PCT_VS_EXCELLENT")) %>%
  mutate(DECIMAL_VALUE = VALUE * 0.01) %>%
  mutate(SERIES_ID = case_when(
    SERIES_ID == "PCT_VS_VERYPOOR" ~ "Very Poor (300–499)",
    SERIES_ID == "PCT_VS_POOR" ~ "Poor (500–600)",
    SERIES_ID == "PCT_VS_FAIR" ~ "Fair (601–660)",
    SERIES_ID == "PCT_VS_GOOD" ~ "Good (661–780)",
    SERIES_ID == "PCT_VS_EXCELLENT" ~ "Excellent (781–850)",
    TRUE ~ SERIES_ID
  )) %>%
  mutate(SERIES_ID = fct_relevel(
    SERIES_ID,
    "Excellent (781–850)",
    "Good (661–780)",
    "Fair (601–660)",
    "Poor (500–600)",
    "Very Poor (300–499)"
  ))

credit_score_plot <- ggplot(
  mortgages_credit_score,
  aes(x = YEAR, y = DECIMAL_VALUE, fill = SERIES_ID)
) +
  geom_area(alpha = 0.75, color = "black", linewidth = 0.5) +
  
  # Vertical reference lines
  #geom_vline(xintercept = 2010, linetype = "dashed", linewidth = 0.4, alpha = 0.6) +
  #geom_vline(xintercept = 2014, linetype = "dashed", linewidth = 0.4, alpha = 0.6) +
  
  geom_segment(aes(x = 2010, xend = 2010, y = 0, yend = 0.99),
             linetype = "dashed", linewidth = 0.4, alpha = 0.4,
             inherit.aes = FALSE) +
  
  geom_segment(aes(x = 2014, xend = 2014, y = 0, yend = 0.99),
             linetype = "dashed", linewidth = 0.4, alpha = 0.4,
             inherit.aes = FALSE) +
  
  # Labels at the top
  annotate("text",
           x = 2010, y = 0.88,
           label = "Dodd-Frank",
           angle = 90,
           hjust = 0.5, vjust = -0.5,
           size = 2.5) +
  annotate("text",
           x = 2014, y = 0.91,
           label = "ATR/QM",
           angle = 90,
           hjust = 0.5, vjust = -0.5,
           size = 2.5) +
  
  labs(
    x = "\nYear",
    y = "Percentage of Mortgage Originations\n",
    fill = "Credit Score"
  ) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(labels = percent) +
  theme_minimal()

print(credit_score_plot)

ggsave("../plots/stacked_area_chart_share_of_credit_rated_mortgages.png", width = 7, height = 3.8, dpi = 300)
```

Figure 5 shows a stacked area chart of mortgage originations based on proportion of credit scores. Prior to Dodd-Frank, risky mortgages (Very Poor, Poor, and Fair) made up 30.0%, on average, of mortgage originations from 1998-2010, 11.8% from 2010-2014, and 15.9% from 2014-2021. The trend remains clear that lenders will take on borrowers with a Good or Excellent credit rating and are hesitant to originate mortgages for a Fair credit score, especially for Poor and Very Poor scores, which the latter has never made up more than 0.20% of all mortgages since 2010.

# III. The Increasing Shift in the Age of Borrowers

This section assesses the age of borrowers from 1998-2021, which are split up into six categories based on NMDB's classification: Under 25, 25-34, 35-44, 45-54, 55-64, 65 & Over.

```{r table_age, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE}

age_table <- data.frame(
  "Age of Borrower" = c("65 & Over", "55-64", "45-54", "35-44", "25-34", "Under 25"),
  "1998" = c("4.3%", "9.9%", "24.0%", "34.1%", "24.8%", "2.9%"),
  "2021" = c("11.5%", "17.0%", "22.9%", "26.7%", "19.8%", "2.1%"),
  "Rate of Change" = c("167.4%", "71.7%", "-4.6%", "-21.7%", "-20.2%", "-27.6%"),
  check.names = FALSE
)

rate_numeric <- as.numeric(gsub("%","", age_table$`Rate of Change`))
top_two <- rank(-rate_numeric) <= 2

age_table$`Rate of Change` <- cell_spec(
  age_table$`Rate of Change`,
  background = ifelse(rate_numeric < 0, "#f8d7da",   # light red
               ifelse(top_two, "#d4edda", "white"))  # light green
)

kable(
  age_table,
  caption = "Rate of change in borrower age categories from 1998 to 2021.",
  escape = FALSE,
  booktabs = TRUE,
  align = "c"
)
```

As shown in Table 1, there were only two groups that experienced a positive rate of change from 1998 to 2021: 55-64 and 65 & Over, with rates of 71.7% and 167.4%, respectively. Despite these drastically large changes, they only represent the fourth and fifth largest groups out of all borrowers in 2021: 17.0% for 55-64 and 11.5% for 65 & Over. Both groups held the same positions in 1998 but at a much lower rate: 9.9% for 55-64 and 4.3% for 65 & Over. The largest age category in 2021 was 35-44, at 26.7%, which was also the largest group in 1998, at 34.1%.

So, while the age of the primary borrower has shifted over time, reflective of a longer living population, the prominent borrower has remained in the 35-44 age category. This trend is shown in Figure 4 with a stacked bar plot.

```{r age_plot, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=6, fig.height=3.8, fig.cap="Stacked bar plot of borrower age categories based on percentage from 1998–2021.", fig.pos='H'}
# STACKED BAR PLOT: Borrower age

mortgages_age_percentages <- mortgages %>%
  filter((MARKET == "All Mortgages") &
           (SERIES_ID == "PCT_AGE_LT25" |
              SERIES_ID == "PCT_AGE_2534" |
              SERIES_ID == "PCT_AGE_3544" |
              SERIES_ID == "PCT_AGE_4554" |
              SERIES_ID == "PCT_AGE_5564" |
              SERIES_ID == "PCT_AGE_GE65")) %>%
  mutate(DECIMAL_VALUE = VALUE * 0.01) %>%
  mutate(SERIES_ID = case_when(
    SERIES_ID == "PCT_AGE_LT25" ~ "Under 25",
    SERIES_ID == "PCT_AGE_2534" ~ "25-34",
    SERIES_ID == "PCT_AGE_3544" ~ "35-44",
    SERIES_ID == "PCT_AGE_4554" ~ "45-54",
    SERIES_ID == "PCT_AGE_5564" ~ "55-64",
    SERIES_ID == "PCT_AGE_GE65" ~ "65 & Over",
    TRUE ~ SERIES_ID
  ))

mortgages_age_percentages$SERIES_ID <- factor(
  mortgages_age_percentages$SERIES_ID,
  levels = c("65 & Over", "55-64", "45-54", "35-44", "25-34", "Under 25"),
  ordered = TRUE
)

borrowers_age_plot <- ggplot(mortgages_age_percentages, aes(x = YEAR, y = DECIMAL_VALUE, fill = SERIES_ID)) +
  geom_bar(stat = "identity", position = "stack", col = "black", linewidth = 0.5, width = 1) +
  labs(
    x = "\nYear",
    y = "Percentage of All Mortgage Borrowers\n",
    fill = "Age of Borrower"
  ) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(borrowers_age_plot)
```

# IV. Changes in Gender & Number of Borrowers Per Mortgage

NMDB categorizes the number of borrowers into three categories: Single Borrower, Two Borrowers, and Multiple Borrowers; but divides Single Borrower by gender, so there are four categories total: Single Borrower - Male, Single Borrower - Female, Two Borrowers, and Multiple Borrowers. Gender information isn't provided for the last two categories, so while Two Borrowers is assumed to be a heterogeneous couple, it is best not to make no assumptions; nor can any assumptions be made for Multiple Borrowers since the number of borrowers would need to be known before making any assumptions, which would then be based on chance.

Two Borrowers were the dominant borrower type in 1998, making up 64.3% of all mortgage originations, while Single Borrower - Male accounted for 18.9% and Single Borrower - Female 16.2% (Figure 5). By the end of 2021, Two Borrowers significantly decreased to 44.4%, while Single Borrower - Male and Single Borrower - Female increased to 33.5% and 21.4%, respectively. Note that Multiple Borrowers held the same percentage of 0.6% in 1998 and 2021.

```{r pie_charts_borrowers, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=5, fig.height=6.5, fig.cap="Distribution of mortgage borrowers by gender and number of borrowers in 1998 and 2021.", fig.pos='H'}
# Was 6 x 8

# Create decimal values for 1998 pie chart
mortgages_borrowers_gender_1998 <- data.frame(
  SERIES_ID = c(
    "Two Borrowers",
    "Single Borrower - Male",
    "Single Borrower - Female",
    "Multiple Borrowers"
  ),
  DECIMAL_VALUE = c(0.643, 0.189, 0.162, 0.006)
)

mortgages_borrowers_gender_1998$label <- ifelse(
  mortgages_borrowers_gender_1998$SERIES_ID == "Multiple Borrowers",
  "",
  paste0(round(mortgages_borrowers_gender_1998$DECIMAL_VALUE * 100, 1), "%")
)

# Create decimal values for 2021 pie chart
mortgages_borrowers_gender_2021 <- data.frame(
  SERIES_ID = c(
    "Two Borrowers",
    "Single Borrower - Male",
    "Single Borrower - Female",
    "Multiple Borrowers"
  ),
  DECIMAL_VALUE = c(0.444, 0.335, 0.214, 0.006)
)

mortgages_borrowers_gender_2021$label <- ifelse(
  mortgages_borrowers_gender_2021$SERIES_ID == "Multiple Borrowers",
  "",
  paste0(round(mortgages_borrowers_gender_2021$DECIMAL_VALUE * 100, 1), "%")
)

# Keep legend order consistent
borrower_order <- c(
  "Single Borrower - Male",
  "Single Borrower - Female",
  "Two Borrowers",
  "Multiple Borrowers"
)

mortgages_borrowers_gender_1998$SERIES_ID <- factor(
  mortgages_borrowers_gender_1998$SERIES_ID,
  levels = borrower_order
)

mortgages_borrowers_gender_2021$SERIES_ID <- factor(
  mortgages_borrowers_gender_2021$SERIES_ID,
  levels = borrower_order
)

# 1998 pie chart
p1998 <- ggplot(mortgages_borrowers_gender_1998,
                aes(x = "", y = DECIMAL_VALUE, fill = SERIES_ID, width = 0.2)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = label),
            position = position_fill(vjust = 0.5),
            size = 3,
            fontface = "bold") +
  coord_polar(theta = "y") +
  labs(
    x = NULL,
    y = NULL,
    fill = "Number of Borrowers"
  ) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  ggtitle("1998") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  guides(fill = guide_legend(reverse = FALSE))

# 2021 pie chart
p2021 <- ggplot(mortgages_borrowers_gender_2021,
                aes(x = "", y = DECIMAL_VALUE, fill = SERIES_ID, width = 0.2)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = label),
            position = position_fill(vjust = 0.5),
            size = 3,
            fontface = "bold") +
  coord_polar(theta = "y") +
  labs(
    x = NULL,
    y = NULL,
    fill = "Number of Borrowers"
  ) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  ggtitle("2021") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  guides(fill = guide_legend(reverse = FALSE))

# Display both pie charts together
#grid.arrange(p1998, p2021, ncol = 1)

combined_pies <- arrangeGrob(p1998, p2021, ncol = 1)

grid.arrange(p1998, p2021, ncol = 1)

ggsave(
  "../plots/pie_charts_gender_no_of_borrowers.png",
  plot = combined_pies,
  width = 5,
  height = 8,
  dpi = 300
)
```

If both Single Borrower categories were combined into one category, it would have become the dominant borrower type in 2021, totaling 54.9%, a large increase from the combined total of 35.1% in 1998. While there are more single borrowers than ever before, this doesn't mean they are more single people—in the romantic sense—than ever before. A spouse or partner may have a bad credit score so one partner signs a mortgage alone; an individual investor may want to purchase a home to generate income, such as with Airbnb or becoming a landlord; or a romantically single person may acquire a mortgage since it's more affordable to own a home than rent a dwelling. Regardless of who the single borrower is and why or how they acquired a mortgage, it appears that this combined group will remain the dominant borrower for some time to come.

# V. Conclusion

This cursory report has identified key economic events and legislation that have affected mortgages from 1998-2021. Prior to the subprime mortgage crisis of 2007-2009, lenders originated about 30.0% of mortgages for borrowers with a Poor credit score or below, but after the Dodd-Frank Act was passed in 2010, lenders originated about 14.5% of these same rated mortgages, ending 2021 with 11.1% of said mortgages. Borrowers, therefore, are required to have more sound credit scores now than ever before.

Due to systemic and demographic changes, the typical borrower profile in 2021 consists of the following characteristics: they will most likely be 35-44, but they have the same chance of being 55-64 as much as being 25-34; they have a greater chance of being an individual rather than a dual borrower; they will have a good or excellent credit score; and they will be investment minded, possibly acquiring a home to generate income, and will almost certainly refinance a new home mortgage if they see interest rates falling to take advantage of cost savings. So, while financial regulations and lender practices have changed over twenty three years, some borrower habits, such as finding ways to save money, have not change and probably never will.
